\morus: Biases in the Stream

### 2018-05-10 morusAC_00_Abstract.tex
Proofreading

% More precisely, we show that a linear correlation exists between some bits of the plaintext and ciphertext. Moreover the bias depends purely on the plaintext, and not on the secret key nor the nonce, hence restricting the single use of nonce or even rekeying does not prevent the attacks.
%
% From the nature of the linear cryptanalysis, the attack works in the known plaintext attacks. Moreover, if a fixed plaintext is encrypted multiple times, the correlation can be used to recover a 1-bit information of the plaintext. The attack principally has larger impact because it can be used to recover more unknown bits of the plaintext, provided an initial segment of the plaintext is known. 
%
  To identify the bias, we exploit rotational symmetries in \morus
  using linear masks that are invariant by word-rotations of the
  state.  This motivates us to introduce word-oriented versions of
  \morus called \minimorus, which help the
  analysis. The attack has been implemented and verified on
  \minimorus, where it yields a bias of $2^{-16}$.

@@

  We also study reduced versions of the initialization and
  finalization of \morus to evaluate the security margin of these
  components.  We show a forgery attack when the finalization is
  reduced from 10 rounds to 3, and a key-recovery attack in the
  nonce-misue setting when the initialization is reduced from 16 to 10
  steps.  These additional results do not threaten full \morus, but
  studying all aspects of the design is useful to understand its
  strengths and weaknesses.

>>>>

% More precisely, we show that a linear correlation exists between some bits of the plaintext and ciphertext. Moreover the bias depends purely on the plaintext, and not on the secret key nor the nonce, hence restricting the single use of nonce or even rekeying does not prevent the attacks.
%
% From the nature of the linear cryptanalysis, the attack works in the known plaintext attacks. Moreover, if a fixed plaintext is encrypted multiple times, the correlation can be used to recover a 1-bit information of the plaintext. The attack principally has larger impact because it can be used to recover more unknown bits of the plaintext, provided an initial segment of the plaintext is known. 
%
  To identify this bias, we exploit rotational symmetries in \morus
  using linear masks that are invariant by word-rotations of the
  state.  This motivates us to introduce single word versions of
  \morus called \minimorus, which help the
  analysis. The attack has been implemented and verified on
  \minimorus, where it yields a bias of $2^{-16}$.

@@

  We also study reduced versions of the initialization and
  finalization of \morus to evaluate the security margin of these
  components.  We show a forgery attack when the finalization is
  reduced from 10 rounds to 3, and a key-recovery attack in the
  nonce-misue setting when the initialization is reduced from 16 to 10
  steps.  These additional results do not threaten the full \morus, but
  studying all aspects of the design is useful to understand its
  strengths and weaknesses.

### 2018-05-10 morusAC_01_Introduction.tex
Proofreading

Our main result is a linear approximation \cite{eurocryptMatsui93,eurocryptMatsuiY92} linking ciphertext bits of five consecutive encryption blocks.
%As a consequence of this trail, we exhibit a biased linear combination of ciphertext bits, which holds across any given sequence of five consecutive encryption steps.
%Technically, this assumes that the inner state of the cipher at the start of the trail is uniformly random, but this is a fair assumption since the inner state should be indistinguishable from uniform randomness by design.
%Moreover, that is the only assumption required; in particular, the secret key or the nonce used to initialize the inner state of MORUS are irrelevant: the bias is independent of their value.
More precisely, we show that when encrypting a fixed plaintext block multiple times, a linear correlation exists between some output bits of the cipher.
Moreover, the bias depends purely on the plaintext, and not on the secret key of the cipher.
In principle, this property could be used to recover unknown bits of a plaintext encrypted a large number of times, provided an initial segment of the plaintext is known.
For \morus[1280] with 256-bit keys, the linear bias is $2^{-73}$
\todo{double-check bias vs. correlation everywhere}
and can be exploited using about $2^{152}$ encrypted blocks.
\todo{double-check}
We note that rekeying does not prevent the attack: the biases are independent of the secret encryption key, and can be exploited as long as a given plaintext block is encrypted sufficiently often, regardless of whether each encryption uses a different key.

>>>>

Our main result is a linear approximation \cite{eurocryptMatsui93,eurocryptMatsuiY92} linking ciphertext bits of five consecutive encryption blocks.
%As a consequence of this trail, we exhibit a biased linear combination of ciphertext bits, which holds across any given sequence of five consecutive encryption steps.
%Technically, this assumes that the inner state of the cipher at the start of the trail is uniformly random, but this is a fair assumption since the inner state should be indistinguishable from uniform randomness by design.
%Moreover, that is the only assumption required; in particular, the secret key or the nonce used to initialize the inner state of MORUS are irrelevant: the bias is independent of their value.
More precisely, we show that when encrypting a fixed plaintext block multiple times, a linear correlation exists between some output bits of the cipher.
Moreover, the bias depends purely on the plaintext, and not on the secret key of the cipher.
In principle, this property could be used to recover unknown bits of a plaintext encrypted a large number of times, provided an initial segment of the plaintext is known.
For \morus[1280] with 256-bit keys, the linear bias is $2^{-73}$
\todo{double-check bias vs. correlation everywhere}
and can be exploited using about $2^{152}$ encrypted blocks.
\todo{double-check}
We note that our attack is nonce-respecting, and that rekeying does not prevent it: the biases are independent of the secret encryption key and nonce, and can be exploited as long as a given plaintext block is encrypted sufficiently often, regardless of whether each encryption uses a different key.
